---
title: "R Tutorial: Answers to tasks"
output: 
  html_document:
    includes:
      before_body: header.htm
---


# Part 1

> **Task**: can you apply the functions `sum` and `length` to boolean vectors?

Yes - sum counts the number of `TRUE` values.


> **Task**: make a list of booleans where the items are `TRUE` if each corresponding item of `nums` is less or equal to 3

```{r eval=F}
nums <= 3
```

> **Task**: make a list of booleans where the items are `TRUE` if each corresponding item of `nums` is either 5 or 1.

```{r eval=F}
nums %in% c(5,1)
```

> **Task **: Using y to index `nums` worked because the length of the two vectors are the same.  What happens if you make another variable `y2` which only has 3 boolean items, and try to index `nums` using that?

```{r eval=F}
y2 = c(TRUE, FALSE, TRUE)
nums[y2]
```


> **Task**:  Find the favourite numbers of all my friends who have more than 4 characters in their name.
> 
> Hint: build this up step by step.  First of all, get the number of characters in each name, then test whether this number is greater than 4.  This should result in a vector of booleans.  Then index nums using this vector.

```{r eval=F}
nums[nchar(friends)>4]
```


# Part 3

> **Task**: What happens if you type `d` into the console to see what's inside the object `d`?

You get a lot of data printed to the screen!  Very unhelpful.


> **Task**: Make a boolean vector which is `TRUE` if the `focal_year` variable is greater than 1850, and `FALSE` otherwise.  Assign this to a variable named `recentFocalYears` (create a variable called `recentFocalYears` and store the boolean vector inside it).

```{r eval=F}
recentFocalYears <- d$focal_year > 1850
```

> **Task**: Make a table of counts of roof shape types for languages in the recent data.  *Hint*: you should index the rows with the variable recentFocalYears and the **code_label** column.

```{r eval=F}
table(d[recentFocalYears,]$code_label)
```

> **Task**: Load the data in `data/Glottolog_Data.csv` into a data frame called glottoData.

```{r eval=FALSE }
rainfallData <- read.csv("../data/RainfallData.csv", stringsAsFactors = F)
```

> **Task** What are the names and formats of the variables in rainfallData? What is the name of the column that includes the numeric amount of rainfall?

```{r eval=FALSE}
str(rainfallData)
```

The `code` variable holds the data on rainfall)




# Part 4

> **Task**: Add a line to your script to make a variable in `d` named `flatRoof` that is TRUE if the roof type is "Flat" and FALSE otherwise.

```{r eval=F}
d$flatRoof <- d$roofShape == "Flat"
```

> **Task**:  Which basic word order type has the higest mean rainfall? Add this to your script


```{r eval=F}
rainfallByLangFam.MEAN <- tapply(d$rainfall, d$language_family, mean)
rainfallByLangFam.MEAN.sorted <- sort(rainfallByLangFam.MEAN, decreasing = TRUE)
head(rainfallByLangFam.MEAN.sorted)
```

> **Task**:  Make a boxplot showing the mean rainfall for each type of roof shape. There are lots of types, which makes the legend difficult to read. Can you make a boxplot for just the top 4 most common roof shape types?

```{r eval=F}
# Plot of all the roof shape types:
boxplot(rainfall ~ roofShape, data = d)

# Plot the 5 most common roof shapes:
# Get a table of counts for roof shape:
roofShapeCounts = table(d$roofShape)
# Sort the list and take the top 4
top4RoofShapes = sort(roofShapeCounts,decreasing = T)[1:4]
# This is a list of numbers, but we want the NAMES of this list:
top4RoofShapes = names(top4RoofShapes)
# Plot the data:
boxplot(rainfall ~ roofShape, data = d[d$roofShape %in% top4RoofShapes,])
```


> **Task**:  Change the colour of the line in the `abline` function to green.

```{r eval=F}
abline(h = 0, col = 'green')
```


# Part 5

> **Task**: What happens if you try to calculate the sum of this variable?

You get NA returned!

> **Task** What happens when you try to work out the mean rainfall in the whole dataset?

```{r eval=F}
mean(rainfallData$code)
```

You get an NA value!

> **Task** Which languages have the rainfall data missing?

Here's how to get rows where `code` is NA:

```{r eval=F}
rainfallData[is.na(rainfallData$code),]
```


> **Task**: How many datapoints in `rainfallData$language_family` are missing? Remember that `is.na` creates a vector of booleans and `sum` can count the number of `TRUE` values in a vector.

```{r eval=F}
sum(is.na(rainfallData$language_family))
sum(is.na(rainfallData$code))
```


>  **Task**: Work out the mean rainfall for each language family in the data. There are a few ways to do this, and none is "better" than any other, as long as the code is clear.

If there were no NA values, then we could use a tapply function to solve this:

```{r eval=F}
tapply(rainfallData$code, rainfallData$language_family, mean)
```

But this will produce some NA values.

There are a few ways to solve this. The first is to make a data frame with only rows where the `code` variable is NOT NA:

```{r eval=F}
rainfallData2 = rainfallData[!is.na(rainfallData$code),]
tapply(rainfallData2$code, rainfallData2$language_family, mean)
```

You could also just pass the argument `na.rm=TRUE` as an extra argument to tapply. tapply passes this argument on to the function it's given. Now, `mean()` will ignore any NA values.

```{r eval=F}
tapply(rainfallData$code, rainfallData$language_family, mean, an.rm=TRUE)
```

